---
title: "Top and Bottom Growth Occupations"
author: "George Self"
date: 2018-09-25T08:00:00-07:00
description: "-- The top and bottom growing occupations in Cochise County --"
categories: ["workforce"]
tags: ["employment","growth"]
---

```{r setup, include=FALSE}
# Load some libraries
library(tidyverse)
library(here)
library(DT)
library(widgetframe)

# Set the rproj home directory
here::here()
```


```{r createDFs, echo=FALSE, message=FALSE, warning=FALSE}

# Create two df for the data tables
topjobs <- head(read_csv("occupations.csv"),28)
bottomjobs <- tail(read_csv("occupations.csv"),25) 

```

I’m continuing my exploration of EMSI occupational data and wanted to share another tidbit, the top and bottom occupations for Cochise County by the growth in job postings between 2013 and 2017. For example, "Grounds Maintenance Workers" had 328 job postings in 2013 and 456 in 2017 for an increase of 39%. Here are the top growth jobs (sorted by the percent change in postings).

```{r topTable, echo=FALSE, message=FALSE, warning=FALSE}

f1 <- datatable(topjobs, options = list(pageLength = 7))
frameWidget(f1)

```


The following table shows the bottom occupations, those that lost the greatest number of jobs from 2013 to 2017.

```{r bottomTable, echo=FALSE, message=FALSE, warning=FALSE}

f2 <- datatable(bottomjobs, options = list(pageLength = 7))
frameWidget(f2)

```


These are occupations where it would be challenging for a person to earn a living. For example, at the very bottom of the list is agricultural laborers, where 52% of the available positions in 2013 were lost by 2017.

These tables also include the median wage for workers in these occupations in Cochise County. Many of the fastest growing occupations are also only just above minimum wage while many of the occupations with a significant loss in numbers are higher-wage jobs.