---
title: "Completions and Jobs"
author: "George Self"
date: 2018-10-02T08:00:00-07:00
description: "-- Cochise College Completions vs. The Job Market --"
categories: ["workforce"]
tags: ["courses","employment"]
---


```{r setup, include=FALSE}
# Load some libraries
library(tidyverse)
library(here)
library(widgetframe)
library(DT)

# Set the rproj home directory
here::here()
```


```{r createDF, echo=FALSE, message=FALSE, warning=FALSE}

# Create table
df <- read_csv("jobs_completions.csv")

```

The following table is a completion report for Cochise College where various programs are linked to a job analysis. For example, in the field of education, we had 23 completers in 2003 and 147 in 2017, for a 539% increase. There were 372 annual openings at a median hourly salary of $18.21. To be sure, most education jobs require degrees higher than an Associate’s, but this information is at least an indicator of the likelihood of success for our graduating students whether they continue to pursue an education degree or seek work as an education paraprofessional.

```{r topTable, echo=FALSE, message=FALSE, warning=FALSE}

f1 <- datatable(df, options = list(pageLength = 7))
frameWidget(f1)

```
