---
title: "University Transfer Student Characteristics"
author: "George Self"
date: 2018-09-18T08:00:00-07:00
description: "-- Some demographics for our students transferring to the university --"
categories: ["cochise"]
tags: ["transfer","university"]
---

```{r setup, include=FALSE}
# Load some libraries
library(tidyverse)
library(here)
library(kableExtra)

# Set some defaults for Knitr
knitr::opts_chunk$set(
  collapse = TRUE,
  strip.white = TRUE,
  comment = "#>",
  out.width = "65%",
  message=FALSE,
  warnings=FALSE
)
options(knitr.table.format = "html")


# Set the rproj home directory
here::here()
```


The state of Arizona maintains a database that tracks students who transfer from one of our community colleges to one of our publicly-funded universities. That database includes a huge amount of information about transfer students and from time-to-time, I'll share something from that database in my blog. For today, the following table lists some general characteristics about our transfer students over the past five years.

```{r procTable, echo=FALSE, message=FALSE, warning=FALSE}

# First, find all records in a selected discipline
df <- read_csv("transfer.csv") %>%
  replace(., is.na(.), "")

df %>%
  kable("html", align = c("l","r","r","r","r","r","r","r")) %>%
    kable_styling(bootstrap_options = "striped",
                full_width = F,
                position = "center") %>%
    row_spec(row = 1, 
             bold = TRUE,
             color = "#FFFFFF",
             background = "#1e4d99") %>%
    row_spec(row = 2, 
             bold = TRUE, 
             background = "#d2dff4") %>%
    row_spec(row = 12, 
             bold = TRUE, 
             background = "#d2dff4") %>%
    row_spec(row = 15, 
             bold = TRUE, 
             background = "#d2dff4") %>%
    row_spec(row = 20, 
             bold = TRUE, 
             background = "#d2dff4")
                
```

