---
title: "Discipline Success"
author: "George Self"
date: 2018-10-22T08:00:00-07:00
description: "-- How Successful Are Our Graduates At The University --"
categories: ["cochise"]
tags: ["discipline","university","success"]
---


```{r setup, include=FALSE}
# Load some libraries
library(tidyverse)
library(here)
library(DT)
library(widgetframe)

# Set the rproj home directory
here::here()
```


```{r createDFs, echo=FALSE, message=FALSE, warning=FALSE}

# Create two df for the data tables
df <- read_csv("discipline_success.csv")

```

This week I took a look at the "Course Benchmark Report" at the Arizona ASSIST site, where data about students who transfer from our college to any of the public universities is available. This report shows the success rate for our students.

```{r topTable, echo=FALSE, message=FALSE, warning=FALSE}

f1 <- datatable(df, options = list(pageLength = 7, scrollX = 'true'))
frameWidget(f1)

```

I suppose that the first fact that jumped out at me was the number of Computer Science and English students who transfer; no other discipline comes very close to those numbers. The other thing that caught my attention was the generally high success rate for our disciplines; 75% and above. The lone exception to that number is mathematics, but that is generally known to be a challenging discipline regardless of a student's background. Finally, the withdrawal rate is fairly consistent across all disciplines.