---
title: "Art With R"
author: "George Self"
date: 2018-10-26T08:00:00-07:00
description: "-- It Is Nice To Sometimes Just Create Pretty Plots --"
categories: ["r"]
tags: ["ggplot","math","art"]
---

```{r setup, include=FALSE}
# Load some libraries
library(here)
library(ggplot2)

# Set the rproj home directory
here::here()
```


I recently found an interesting tutorial on how to use *ggplot2* in R to create pretty graphs. The key is to define a mathematical function that plots points in some sort of pattern and then use various *ggplot2* features to make that plot look nice. Find the [tutorial at this link](https://www.datacamp.com/projects/62).

The tutorial used the concept of phyllotaxis to create interesting scatter plots. Phyllotaxis is the spiral arrangement of the leaves on a stem and that arrangement can be simulated with a bit of mathematics. Here are two plots I made using the phyllotaxis concept.

```{r image01, fig.width=5, fig.asp=1.0}
# Load ggplot2
library(ggplot2)

# Define the number of points
points <- 1000

# Defining the angle between each point
# The 'Golden Angle' is found in nature, about 2.4
#angle <- pi*(3 - sqrt(5)) # About 2.4
angle <- 3.5 # Tried various angles other than the Golden Angle

# Transparancy for the points (0-1.0)
ptAlpha <- 0.25

# Shape of the points (see http://www.cookbook-r.com/Graphs/Shapes_and_line_types/)
ptShape <- 19

# Set the background color
bgColor <- 'lightyellow'

# Create a df with the points to plot
t <- (1:points) * angle
x <- sin(t)
y <-cos(t)
df <- data.frame(t, x, y)

ggplot(df, aes(x*t, y*t)) +
  geom_point(aes(color = t), size = 10.0, alpha = ptAlpha, shape = ptShape) +
  scale_color_gradient(low = 'orangered1', high = 'gold2') +
  theme(panel.background = element_rect(fill = bgColor),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

```


```{r image02, fig.width=5, fig.asp=1.0}
# These lines are explained in the first image

library(ggplot2)
points <- 500
angle <- 7.0
ptColor <- 'darkgreen'
ptAlpha <- 0.5
ptShape <- 8
bgColor <- '#eeffee'

t <- (1:points) * angle
x <- sin(t)
y <-cos(t)
df <- data.frame(t, x, y)

ggplot(df, aes(x*t, y*t)) +
  geom_point(aes(size = t, color = t), alpha = ptAlpha, shape = ptShape) +
  scale_color_gradient(low = 'blue4', high = 'green4') +
  theme(panel.background = element_rect(fill = bgColor),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

```



The mathematics for these images was fairly simple and it took me longer to figure out all of the *ggplot2* settings needed to make the plots look nice without all of the analysis trappings (like gridlines). I had a couple of hours of fun creating this post --- now, back to work...
